---
title: "MSDS 6306:  Doing Data Science - Tidy Data"
subtitle: "Live session Unit 05 assignment"
author: "Selwyn Samuel"
date: "June 11, 2018"
output: 
  html_document:
     fig_caption: yes
---

#### **Due: 1 hour before your 6^th^ live session**
$~$

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### **Submission**
**ALL MATERIAL MUST BE KNITTED INTO A SINGLE, LEGIBLE, AND DOCUMENTED HTML DOCUMENT. Formatting can be basic, but it should be easily human-readable.  Unless otherwise stated, please enable {r, echo=TRUE} so your code is visible.**
$~$

#### **Questions**

**Backstory:** Your client is expecting a baby soon.  However, he is not sure what to name the child.  Being out of the loop, he hires you to help him figure out popular names.  He provides for you raw data in order to help you make a decision.

**1. Data Munging (30 points): Utilize yob2016.txt for this question. This file is a series of popular children's names born in the year 2016 in the United States.  It consists of three columns with a first name, a gender, and the amount of children given that name.  However, the data is raw and will need cleaning to make it tidy and usable.**
$~$

**1A. First, import the .txt file into R so you can process it.  Keep in mind this is not a CSV file.  You might have to open the file to see what you're dealing with.  Assign the resulting data frame to an object, df, that consists of three columns with human-readable column names for each.**

```{r import_yob2016.txt}
#
# Read file into a data frame. The file has ";" delimiter. So we need to use read.table instead of read.csv.
#
df <- read.table("C:/SMU/SMU-MSDS-Homework/6306-Unit5/yob2016.txt", header=FALSE, sep=";")

#
# Assign human-readable column names 
#
names(df) <- c("FirstName", "Gender", "Count2016")
```
$~$

**1B. Display the summary and structure of df.**
```{r summary_yob2016.txt}

#
# Summary
#
summary(df)

#
# Structure
#
str(df)

#
# Dimensions
#
dim(df)
```
$~$

**1C  Your client tells you that there is a problem with the raw file.  One name was entered twice and misspelled.  The client cannot remember which name it is; there are thousands he saw! But he did mention he accidentally put three y's at the end of the name.  Write an R command to figure out which name it is and display it.**

```{r munge_yob2016.txt}
#
# Find the name ending with three y's and display it. Get the row index and store it in a variable for later use.
#
index <- grep("yyy$", df$FirstName)
df[index,]
```
$~$

**1D. Upon finding the misspelled name, please remove this particular observation, as the client says it's redundant.  Save the remaining dataset as an object: y2016.**
```{r remove_name_yob2016.txt}
#
# We have the row number in the index variable from previous step
#
y2016 <- df[-c(index),]

#
# Display dimensions and do a search for three y's to make sure that row was removed.
#
dim(y2016)
grep("yyy$", y2016$FirstName)
```
$~$

**2. Data Merging (30 points): Utilize yob2015.txt for this question.  This file is similar to yob2016, but contains names, gender, and total children given that name for the year 2015.**

**2A. Like 1a, please import the .txt file into R.  Look at the file before you do.  You might have to change some options to import it properly.  Again, please give the dataframe human-readable column names.  Assign the dataframe to y2015.**

```{r import_yob2015.txt}
#
# Read file into a data frame. The file has "," delimiter. So we can use read.csv.
#
y2015 <- read.csv("C:/SMU/SMU-MSDS-Homework/6306-Unit5/yob2015.txt", header=FALSE)

#
# Assign human-readable column names 
#
names(y2015) <- c("FirstName", "Gender", "Count2015")
```
$~$

**2B. Display the last ten rows in the dataframe.  Describe something you find interesting about these 10 rows.**

```{r display_yob2015.txt}
#
# Display last 10 rows.
#
tail(y2015, 10)
```

All the last rows start with the letter 'Z', male names and all of them were used exactly by 5 babies.
$~$

**2C. Merge y2016 and y2015 by your Name column; assign it to final.  The client only cares about names that have data for both 2016 and 2015; there should be no NA values in either of your amount of children rows after merging.**

```{r data_merge}
#
# Merge y2015 & y2016 by the firstname and assign to final
#
final <- merge(y2015, y2016, by="FirstName")
head(final)

#final[final$Gender.x != final$Gender.y,]

#
# Print row counts
#
message("y2015: ", nrow(y2015))
message("y2016: ", nrow(y2016))
message("final: ", nrow(final))

#
# Check for NA in any of the amount columns
#
any(is.na(final$Count2015))
any(is.na(final$Count2016))
```
$~$

**3. Data Summary (30 points): Utilize your data frame object final for this part.**

**3A. Create a new column called "Total" in final that adds the amount of children in 2015 and 2016 together.  In those two years combined, how many people were given popular names?**

```{r data_summary}
#
# Add a new column that has the sum of Count2015 and Count2016
#
final$Total <- final$Count2015 + final$Count2016
```
$~$

**3B. Sort the data by Total.  What are the top 10 most popular names?**
```{r data_sort}
#
# Sort the data frame descending based on the Total column
#
sorted_final <- final[order(-final$Total),]

#
# List the top 10 popular names
#
head(sorted_final, 10)
```
$~$

**3C. The client is expecting a girl!  Omit boys and give the top 10 most popular girl's names.**
```{r data_top10_girls_names}
#
# get the top 10 girls names
#
sorted_girls <- sorted_final[sorted_final$Gender.x == 'F' | sorted_final$Gender.y == 'F',]
head(sorted_girls, 10)
```
$~$

**3D. Write these top 10 girl names and their Totals to a CSV file.  Leave out the other columns entirely.**
```{r write_top_girlsnames_csv}
#
# get top 10 girls names to a variable
#
top10_girls <- head(sorted_girls, 10)[, c("FirstName", "Total")]
write.csv(top10_girls, "C:/SMU/SMU-MSDS-Homework/6306-Unit5/top10girls.csv", row.names = FALSE)
```

**4. Upload to GitHub (10 points): Push at minimum your RMarkdown for this homework assignment and a Codebook to one of your GitHub repositories (you might place this in a Homework repo like last week).  The Codebook should contain a short definition of each object you create, and if creating multiple files, which file it is contained in.  You are welcome and encouraged to add other files-just make sure you have a description and directions that are helpful for the grader.**

```{r generate_codebooks}
#
# Ideas From: https://www.r-bloggers.com/generating-codebooks-in-r/
# Link shared by: Shon Mohsin in slack channel
#
library(dataMaid)

#
# We will generate code books for "final" and "top_10_girls" objects.
#

# Set the descriptions for each object and the contents
attr(final, "labels") <- "List of baby names from the years 2015 & 2016 along with the gender and the total count of the number of times the baby name was used."
attr(final$FirstName, "shortDescription") <- "First name of the baby"
attr(final$Gender.x, "shortDescription") <- "Gender of the baby in 2015 report"
attr(final$Count2015, "shortDescription") <- "Count of the baby name in 2015"
attr(final$Gender.y, "shortDescription") <- "Gender of the baby in 2016 report"
attr(final$Count2016, "shortDescription") <- "Count of the baby name in 2016"
attr(final$Total, "shortDescription") <- "Total count of the baby name in 2015 & 2016"
makeCodebook(final, reportTitle = "Data frame for baby names in 2015 & 2016")

attr(top10_girls, "labels") <- "List of top 10 girl baby names from the years 2015 & 2016 and the total count of the number of times the baby name was used."
attr(top10_girls$FirstName, "shortDescription") <- "First name of the baby"
attr(top10_girls$Total, "shortDescription") <- "Total count of the baby name in 2015 & 2016"
makeCodebook(top10_girls, reportTitle = "Data frame for top 10 girl baby names in 2015 & 2016")

```
$~$

Link: https://github.com/selwyn-samuel-smu/SMU-MSDS-Homework/tree/master/6306-Unit5

$~$

**Reminder**

**To complete this assignment, please submit one RMarkdown and matching HTML file at least one hour before your live session on June 14, 2018.  You do not need to submit a link to your GitHub: just note where the assignment is (URL) in your RMarkdown file.  Make sure it is public!! Please submit all files at the same time; only one submission is granted.**
